---
title: "densr"
output: 
  bookdown::html_document2:
      base_format: rmarkdown::html_vignette
      fig_caption: yes
      number_sections: no
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{densr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bookdown}
  %\VignetteDepends{IMIFA}
  %\VignetteDepends{MoEClust}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(densr)
```

## Introduction

Welcome to the __densr__ package, a powerful tool designed to enhance your data analysis by providing methods to overlay histograms with various density estimation techniques. This vignette will guide you through the installation process and demonstrate the usage of the central components of this package.

Three functions are provided:

  1. `frequency_polygon()`: for calculating a frequency polygon density estimate,
  
  2. `create_hist_dens()`: for creating objects of class hist_dens,
  
  3. and a `plot()` method for producing final visualisations.
  
Typically, these functions are applied sequentially (i.e. used in the order that they are listed above). See below for a typical workflow of the `densr` package.

### Installation

__densr__ will run in Windows, Mac OS X, or Linux but firstly requires installing R [R](https://cran.r-project.org/), and R Studio [RStudio](https://www.rstudio.com/) (highly recommended).

Type the following command in R to install the latest development version of the package from the __densr__ [GitHub page](https://github.com/rachel-mc/densr): 

```{r, eval=FALSE}
install.packages('devtools')
devtools::install_github('rachel-mc/densr')
```

Then type:

```{r}
library(densr)
```

to load in the aforementioned functions.

Now that the `densr` package is installed and loaded, let's explore its functionalities.

## `frequency_polygon()`

This function derives a frequency polygon from the components of a histogram object and returns the coordinates of the resulting density estimate. The frequency polygon is a valid density as it integrates to approximately one. This function is used within `create_hist_dens()` if `type = "fp"`.

```{r}
data(ais, package = "MoEClust")
fem <- subset(ais, sex == "female")$Fe
frequency_polygon(fem)
```

## `create_hist_dens()`

This is the main function of the `densr` package as it is used to ultimately create density histograms. It creates an object of class "hist_dens", which contains the data, histogram information (to later use inside the plot function), density values, and the particular method used for estimating the density.

The `type` argument consists of a character string denoting the valid options for different density estimations. `match.arg()` and the intelligible error messages included in the function additionally check that the user correctly supplied _one_ of the following.   If no type is specified, the default is used and otherwise execution is terminated:  
  "gaussian": Gaussian density estimation  
  "kde": Kernel density estimation  
  "fp": Frequency polygon density estimation  
The default type is specified as the first entry in the character string (i.e. "gaussian").

We illustrate this function's default settings with a toy example that generates random values from a standard normal distribution, using the default `rnorm()` arguments of mean 0 and standard deviation 1. Recall that `create_hist_dens(x)` will be equivalent to `create_hist_dens(x, type = "gaussian")` by default.

```{r}
fake_data <- rnorm(10000)
x <- create_hist_dens(fake_data)
```

Kernel density estimation is the most commonly used nonparametric density estimator. The ellipses construct allows additional arguments to the density function to be supplied. 

This example uses Old Faithful Geyser Data and its waiting time between eruptions variable which exhibits multimodality. R estimates the optimal bandwidth (degree of smoothing) when it is not specified by the user. We can manually adjust the kernel as follows:

```{r}
data(faithful)
data1 <- create_hist_dens(faithful$waiting, type = "kde", kernel = "rectangular")
data2 <- create_hist_dens(faithful$waiting, type = "kde", kernel = "biweight")
```

We can also pass additional arguments to the inner `hist()` function. Below, one histogram has the number of intervals chosen by Scott's Rule, and the other has a sequence of breaks defined for it that spans the range of the data.

```{r}
data(olive, package = "IMIFA")
scott <- create_hist_dens(olive$linolenic, type = "fp", breaks = "Scott")
seq <- create_hist_dens(olive$linolenic, type = "fp", breaks = seq(0, 80, by = 2))
```

## `plot()`

To visualize the histograms with a density overlayed, you can use the dedicated plotting function. The `plot()` method provided simply plots the created objects of class "hist_dens". Taking each data from above, a plot can be created by typing `plot()` with the data as its argument.

```{r stdnorm, fig.cap = "Gaussian parametric density estimation overlayed to random draws from a standard normal distribution"}
plot(x, col = "thistle", main = "Standard Normal Distribution")
```

Figure \@ref(fig:stdnorm) indicates that the assumed parametric form for this set of independent and identically distributed observations is correct.

Let's see the impact that changing the kernel has on the density:

```{r fig.cap = "Nonparametric density estimation of waiting times between eruptions of the 'Old Faithful' geyser using a rectangular and biweight kernel respectively"}
par(mfrow = c(1,2))
plot(data1, lwd = 2, main = "", xlab = "Waiting Time (in minutes)")
plot(data2, lwd = 2, lty = 3, main = "", xlab = "Waiting Time (in minutes)")
legend("topleft", legend = c("rectangular", "biweight"), lty = c(1,3), lwd = c(2,2))
```

Next, we compare the two histograms of the same data with different breaks side by side. 

```{r fig.cap = "Frequency Polygon density estimation of alpha-linolenic acid content in olives superimposed to histograms with varying breaks"}
par(mfrow = c(1,2))
suppressWarnings(plot(scott, main = "", border = "darkorchid2", xlab = "Linolenic acid content (%)"))
suppressWarnings(plot(seq, main = "", border = "dodgerblue2", xlab = "Linolenic acid content (%)"))
mtext("Density histograms with Scott's breaks (purple) and Sequence breaks (blue)", side = 1, line = -23, outer = TRUE)
```

### Conclusion

The `densr` package is useful to smoothly combine different methods for estimating density into your histograms. Through trialing and testing various kinds and settings, you can gain a deeper understanding of your data.

If you encounter any issues or have suggestions, please visit the [GitHub issues page](https://github.com/rachel-mc/densr/issues)
